Implicit yaddayadda

### 2011-10-27 papers/PLDI2012/implicitization.tex
quelques TODOs dans le debut

\section{Introduction}
% TODO expliquer 'implicite'
Static analysis by abstract interpretation is a fully automatic program analysis method. When applied to imperative programs, it computes an inductive invariant mapping each program location (or a subset thereof) to a set of states represented symbolically~\cite{CousotCousot_JLC92}.
For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,PPL,BagnaraHZ08SCP}.

@@

It is a known limitation of program analysis by abstract interpretation that this convex hull, or more generally, least upper bound operation, may introduce states that cannot occur in the real program: for instance, the convex hull of the intervals $[-2,-1]$ and $[1,2]$ is $[-2,2]$, strictly larger than the union of the two.
Such introduction may prevent proving desired program properties, for instance $\neq 0$. The alternative is to keep the union symbolic (e.g. compute using $[-2,-1] \cup [1,2]$) and thus compute in the \emph{disjunctive completion}
of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_TASE07,ASTREE_ESOP05,ASTREE_PLDI03}. % TODO ne garder que les papiers les plus significatifs par rapport au sujet évoqué

Assuming we are  interested in a loop-free program fragment, the above approach of keeping symbolic unions gives the same results as performing the analysis separately over every path in the fragment.
A recent method for finding inductive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}.

@@

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
% TODO: 'We implemented' pas une contribution; rephrase.
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
\item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques. (\S\ref{sec:experiments})
\end{enumerate}

@@

@@

The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$. % TODO chercher ref Astree
Let us briefly see why.

Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant. % TODO expliquer widening suivi de narrowing
The \lstinline|x >= x_old+10| test at line~6, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,\allowbreak 100000]$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.

>>>>

\section{Introduction}
Static analysis by abstract interpretation is a fully automatic program analysis method. When applied to imperative programs, it computes an inductive invariant mapping each program location (or a subset thereof) to a set of states represented symbolically~\cite{CousotCousot_JLC92}.
For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,PPL,BagnaraHZ08SCP}.

@@

It is a known limitation of program analysis by abstract interpretation that this convex hull, or more generally, least upper bound operation, may introduce states that cannot occur in the real program: for instance, the convex hull of the intervals $[-2,-1]$ and $[1,2]$ is $[-2,2]$, strictly larger than the union of the two.
Such introduction may prevent proving desired program properties, for instance $\neq 0$. The alternative is to keep the union symbolic (e.g. compute using $[-2,-1] \cup [1,2]$) and thus compute in the \emph{disjunctive completion}
of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_PLDI03,ASTREE_ESOP05}.

Assuming we are  interested in a loop-free program fragment, the above approach of keeping symbolic unions gives the same results as performing the analysis separately over every path in the fragment.
A recent method for finding inductive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}. They can equivalently be seen as analyses over a multigraph of transitions between some distinguished control nodes. This multigraph has an exponential number of edges, but is never explicitly represented in memory; instead, this graph is \emph{implicitly} or \emph{succinctly} represented: its edges are enumerated as needed as solutions to SMT problems.

@@

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
This article summarizes our implementation decisions for analysis of SSA-form programs.
\item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques. (\S\ref{sec:experiments})
\end{enumerate}

@@

Once an inductive invariant $\gamma((\abstr{x}_p)_{p \in P})$ has been obtained, one can attempt \emph{decreasing} or \emph{narrowing} iterations to reduce it. In their simplest form, this just means running the following operation until a fixpoint or a maximal number of iterations are reached: for any $p'$, replace $\abstr{x}_{p'}$ by $\abstr{x}_p \cap \left(\bigsqcup_{p \in P} \abstr{\tau}_{p,p'} (\abstr{x}_p)\right)$. The result also defines an inductive invariant. These decreasing iterations are indispensable to recover properties from guards (tests) in the program in most iteration settings; unfortunately, certain loops, particularly those involving identity (no-operation) transitions, may foil them: the iterations immediately reach a fixpoint and do not decrease further (see example in \S\ref{subsec:rate_lim}). Sections \ref{sec:guided} and \ref{sec:path_focusing} describe techniques that work around this problem.

@@

The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
Let us briefly see why.

Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
The \lstinline|x >= x_old+10| test at line~6, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,\allowbreak 100000]$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.
